\documentclass[11pt]{article}

% --- Encoding & fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% --- Japanese support (compile with lualatex/xelatex, or use CJK with pdflatex) ---
% If you compile with pdflatex, CJKutf8 lets you typeset Japanese inline.
\usepackage{CJKutf8}

% --- Layout & math ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
}

% --- Bibliography ---
\usepackage[backend=biber, style=numeric]{biblatex}
\addbibresource{references.bib}

\title{Japanese Notes --- JLPT N5}
\author{}
\date{\today}

\begin{document}
\begin{CJK}{UTF8}{min}

\maketitle
\tableofcontents

\section{Introduction}
JLPT N5 is the lowest level of the Japanese-Language Proficiency Test. These
notes cover the core hiragana, katakana, basic vocabulary, and grammar needed
to reach this level.

\section{Hiragana}
% Example of typesetting Japanese: あ い う え お

\subsection{Typing small kana (mozc)}
Small kana change the sound and must be typed deliberately.

\textbf{Combined sounds} --- type the whole syllable and the small ゃ/ゅ/ょ
appears automatically:
\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Type & Result \\
\midrule
\texttt{sho} / \texttt{syo} & しょ \\
\texttt{kyo} & きょ \\
\texttt{cho} & ちょ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Small kana on their own} --- prefix with \texttt{x} (or \texttt{l}):
\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Type & Result \\
\midrule
\texttt{xtu} & っ (small tsu --- also made by doubling a consonant, e.g.\ \texttt{kakkoii}) \\
\texttt{xya} & ゃ \\
\texttt{xyu} & ゅ \\
\texttt{xyo} & ょ \\
\texttt{xa} & ぁ \\
\bottomrule
\end{tabular}
\end{center}
Rule of thumb: \texttt{x} (or \texttt{l}) before a kana gives its small version.

\section{Katakana}
Katakana is used mainly for \textbf{loanwords} (words borrowed from other
languages). The long-vowel mark \textbf{ー} (\emph{chōonpu}) marks a lengthened
vowel.
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Katakana & Romaji & Origin & English \\
\midrule
ノート   & nōto   & English ``note''       & notebook \\
コーヒー & kōhī   & English ``coffee''     & coffee \\
テレビ   & terebi & English ``television'' & TV \\
\bottomrule
\end{tabular}
\end{center}
mozc tip: type the word (e.g.\ \texttt{no-to}, where \texttt{-} gives ー) and
press \texttt{F7} to convert to katakana.

Katakana is also commonly used for \textbf{animal and plant names}, even though
they are native Japanese words (a stylistic convention, not loanwords):
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Katakana & Romaji & English \\
\midrule
オウム & oumu  & parrot (as in オウム返し, ``echoing'') \\
イヌ   & inu   & dog \\
ネコ   & neko  & cat \\
サクラ & sakura & cherry blossom \\
\bottomrule
\end{tabular}
\end{center}
These also have standard \textbf{kanji} (猫, 犬, 桜), and the katakana is
optional/stylistic --- the kanji is the most common everyday form. So
猫 = ネコ = ねこ are all the same word (\emph{neko}); either script is fine
(どちらでもいいです).

\section{Vocabulary}

\subsection{Greetings}
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Japanese & Romaji & Type (mozc) & English \\
\midrule
こんにちは & konnichiwa & ko n ni chi ha & Hello / Good afternoon \\
こんばんは & konbanwa   & ko n ba n ha  & Good evening \\
\bottomrule
\end{tabular}
\end{center}

Note: the final \textbf{は} is the topic particle. It is pronounced ``wa'' but
written \textbf{は} and typed as \texttt{ha} in an IME.

\subsection{Time words}
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Japanese & Romaji & English \\
\midrule
今晩 & konban & Tonight / this evening \\
次   & tsugi  & Next \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Adjectives}
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Japanese & Romaji & Type (mozc) & English \\
\midrule
いい     & ii      & i i        & good \\
かっこいい & kakkoii & ka k ko i i & cool / stylish \\
すごい   & sugoi   & su go i    & amazing / great / awesome \\
完璧     & kanpeki & ka n pe ki & perfect / flawless \\
\bottomrule
\end{tabular}
\end{center}
Note the small っ in か\textbf{っ}こいい --- type the doubled consonant
(\texttt{kakkoii}).

\textbf{い-adjectives vs な-adjectives:} いい and かっこいい are
\emph{い-adjectives} (they end in い). 完璧 (かんぺき) is a \emph{な-adjective}:
it takes な before a noun --- 完璧\textbf{な}質問 (``a perfect question'') --- and
完璧です to end a sentence.

\subsection{Nouns}
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Japanese & Romaji & Register & English \\
\midrule
会議       & kaigi   & formal (Sino-Japanese) & meeting / conference \\
ミーティング & mītingu & casual (loanword)      & meeting (team / quick) \\
\bottomrule
\end{tabular}
\end{center}
Both mean ``meeting.'' 会議 is the formal word (official/business); ミーティング
(katakana loanword) is more casual. ``Hold a meeting'' = 会議\textbf{を}します.

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Japanese & Romaji & English \\
\midrule
彼女     & kanojo        & she / girlfriend \\
彼       & kare          & he / boyfriend \\
日本語   & nihongo       & Japanese (language) \\
勉強     & benkyou       & study \\
\bottomrule
\end{tabular}
\end{center}
勉強 is a noun; \textbf{勉強します} (\emph{benkyou shimasu}) = ``to study''
(noun + します). Note the small ょ and long う in べんきょ\textbf{う}, and the small
ょ in かのじょ (彼女).

\section{Grammar}

\subsection{Particle \texorpdfstring{と}{to} (``with'' / ``and'')}
The particle \textbf{と} has two common N5 uses:
\begin{itemize}
  \item ``\textbf{with}'' (a companion): 彼女\textbf{と}日本語を勉強します。 ---
        ``I study Japanese \emph{with} my girlfriend.''
  \item ``\textbf{and}'' (listing nouns): パン\textbf{と}コーヒー --- ``bread
        \emph{and} coffee.''
\end{itemize}
It goes \emph{after} the noun it attaches to (彼女と = ``with her'').

\textbf{Contrast --- と (``and'') vs か (``or''):} between nouns, \textbf{と}
joins them all (``and''), while \textbf{か} offers a choice (``or'').
\begin{itemize}
  \item パン\textbf{と}コーヒー --- ``bread \emph{and} coffee'' (both)
  \item 猫\textbf{か}ネコ --- ``猫 \emph{or} ネコ'' (one of them)
\end{itemize}
``Either is fine'' = どちらでもいいです。

\subsection{Existence: \texorpdfstring{あります}{arimasu} / \texorpdfstring{います}{imasu}}
To say something exists (or someone has something), use the pattern:
\begin{center}
  \textbf{X に　Y が　あります} \quad(\emph{X ni Y ga arimasu})
\end{center}
where \textbf{に} marks the location (where Y is) and \textbf{が} marks the
thing that exists.

\begin{itemize}
  \item あります --- used for \emph{inanimate} things (objects, words, places).
  \item います --- used for \emph{living} things (people, animals).
\end{itemize}

Example:
\begin{quote}
ありがとうに「う」があります。
\end{quote}
\emph{Arigatou ni ``u'' ga arimasu.} --- ``There is an `う' in ありがとう.''

\textbf{Negative and other forms of あります:}
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Form & Japanese & Meaning \\
\midrule
affirmative   & あります             & there is / I have \\
negative      & ありません           & there isn't / I don't have \\
past          & ありました           & there was \\
past negative & ありませんでした     & there wasn't \\
\bottomrule
\end{tabular}
\end{center}
\begin{quote}
時間がありません。 --- ``(I/we) don't have time.'' (lit.\ ``time does not exist'')
\end{quote}

\textbf{General rule:} polite \textbf{〜ます $\rightarrow$ 〜ません} makes any verb
negative --- 行きます $\rightarrow$ 行きません (``don't go''),
わかります $\rightarrow$ わかりません (``don't understand'').

\subsection{Indefinites: question word + \texorpdfstring{か}{ka}}
Adding \textbf{か} to a question word makes it indefinite (``some-/any-''). This
か is \emph{not} the question particle.
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Question word & + か & Romaji & Meaning \\
\midrule
何 (what)   & 何か   & nanika & something / anything \\
誰 (who)    & 誰か   & dareka & someone / anyone \\
どこ (where)& どこか & dokoka & somewhere / anywhere \\
いつ (when) & いつか & itsuka & sometime / someday \\
\bottomrule
\end{tabular}
\end{center}
Example:
\begin{quote}
何か他にありますか。 --- ``Is there anything else?''
\end{quote}
Here there are \emph{two} か: 何\textbf{か} (anything) + the final \textbf{か}
(question particle).

\textbf{か vs が --- different meanings:}
\begin{itemize}
  \item 何\textbf{か}ありますか。--- ``Is there \emph{anything}?'' (yes/no)
  \item 何\textbf{が}ありますか。--- ``\emph{What} is there?'' (asks you to name things)
\end{itemize}

\subsection{Question word: \texorpdfstring{どうやって}{dou yatte}}
\textbf{どうやって} (\emph{dou yatte}) means ``how'' in the sense of
\emph{by what means / method}. It breaks down as:
\begin{itemize}
  \item どう (\emph{dou}) --- how / in what way
  \item やって (\emph{yatte}) --- て-form of やる (\emph{yaru}, ``to do'')
\end{itemize}
Literally ``doing in what way'' $\rightarrow$ ``how / by what means.''

Use it when the answer is a \emph{method} (walking, by train, mixing
ingredients, etc.):
\begin{quote}
駅までどうやって行きますか。 --- ``How do I get to the station?''
\end{quote}

Contrast with plain \textbf{どう}, which asks about a state or impression:
\begin{quote}
どうですか。 --- ``How is it? / How are you?''
\end{quote}

\subsection{Forms of \texorpdfstring{やる}{yaru} (``to do'')}
A single verb branches into several forms. For やる (\emph{yaru}):
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Form & Japanese & Use \\
\midrule
Dictionary  & やる   & ``to do'' \\
て-form      & やって & connector --- ``doing / and do'' (as in どうやって) \\
Casual past & やった & ``did'' --- also the cheer 「やった！」 = ``I did it! / Yay!'' \\
\bottomrule
\end{tabular}
\end{center}
Note that やって and やった differ only in the last kana (て vs た) and both
use the small っ (typed \texttt{yatte} / \texttt{yatta}).

\subsection{Sentence-final particles: \texorpdfstring{よ}{yo} / \texorpdfstring{ね}{ne}}
These attach to the end of a sentence to add nuance:
\begin{itemize}
  \item \textbf{よ} (\emph{yo}) --- emphasis; \emph{informing/asserting} something
        to the listener (``\ldots you know!'').
  \item \textbf{ね} (\emph{ne}) --- seeks \emph{agreement} (``\ldots right?'').
\end{itemize}
\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Sentence & Feel \\
\midrule
かっこいいです。   & ``It's cool.'' (neutral) \\
かっこいいです\textbf{よ}。 & ``It's cool, you know!'' (asserting) \\
かっこいいです\textbf{ね}。 & ``It's cool, isn't it?'' (seeking agreement) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Homophones: \texorpdfstring{起きる}{okiru} vs \texorpdfstring{置く}{oku} (\texorpdfstring{おきます}{okimasu})}
Both verbs become \textbf{おきます} in the polite form --- in kana alone they
are identical, and only the kanji (or context) tells them apart:
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Kanji & Verb & Meaning & Object? \\
\midrule
\textbf{起}きます & 起きる (\emph{okiru}) & to wake up / get up & no (intransitive) \\
\textbf{置}きます & 置く (\emph{oku})    & to put / place      & yes, takes を (transitive) \\
\bottomrule
\end{tabular}
\end{center}
\textbf{Disambiguation trick:} if there is a direct object marked with
\textbf{を}, it is 置く (put). If there is no を (just a time/place), it is
起きる (wake up).
\begin{itemize}
  \item 朝6時に\textbf{起きます}。 --- ``I get up at 6 a.m.'' (no を)
  \item かばん\textbf{を}ここに\textbf{置きます}。 --- ``I put the bag here.'' (を)
\end{itemize}
This is a clear example of why kanji matters: it instantly distinguishes words
that sound identical.

\subsection{Conjugating \texorpdfstring{いい}{ii} (``good'') --- all four forms}
\textbf{いい} (good) has an \emph{irregular} conjugation: in every form except
the plain present, its stem changes to \textbf{よ-}.
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
 & Affirmative & Negative \\
\midrule
present & いい       & よくない \\
past    & よかった   & よくなかった \\
\bottomrule
\end{tabular}
\end{center}
\begin{itemize}
  \item いい --- ``(it's) good''
  \item よくない --- ``(it's) not good''
  \item よかった --- ``(it) was good''
  \item よくなかった --- ``(it) was not good''
\end{itemize}
Never use the いー stem: いかった / いくない ($\times$) are wrong --- always
\textbf{よ-}. 「よかった！」 is also a common reaction (``Oh good! / I'm so glad!'').

\textbf{General rule for い-adjectives:} drop \textbf{い} and add the ending.
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
 & present & negative & past \\
\midrule
高い (expensive) & 高い & 高くない & 高かった \\
\emph{pattern}   & \texttt{-い} & \texttt{-くない} & \texttt{-かった} \\
\bottomrule
\end{tabular}
\end{center}
いい is the one irregular exception (uses よ-). To be polite, add です to any of
these (高くない\textbf{です}); for the past, \emph{never} say でした
(よかった\textbf{です}, not よかったでした $\times$).

\textbf{Verb negatives:} polite \textbf{〜ます $\rightarrow$ 〜ません} (行きます
$\rightarrow$ 行きません); past negative is \textbf{〜ませんでした}
(行きませんでした, ``didn't go'').

\subsection{Demonstratives (\texorpdfstring{こそあど}{kosoado}): \texorpdfstring{これ}{kore} vs \texorpdfstring{この}{kono}}
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
 & Near me (こ) & Near you (そ) & Over there (あ) \\
\midrule
Pronoun (``this/that \textbf{one}'', stands alone) & これ & それ & あれ \\
+ noun (``this/that \textbf{<noun>}'')             & この & その & あの \\
\bottomrule
\end{tabular}
\end{center}
\textbf{Key:} これ stands alone; この must be followed by a noun.
\begin{itemize}
  \item これはよかったです。 --- ``This one was good.'' (これ alone)
  \item この単語はよかったです。 --- ``This word was good.'' (この + noun 単語)
\end{itemize}

There is also a ``\textbf{such / what kind of}'' series. Like この, these attach
\emph{directly} to a noun (no の):
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Word & Meaning & Example \\
\midrule
こんな & this kind of           & こんな本 (this kind of book) \\
そんな & that kind of / such    & そんなイベント (such an event) \\
あんな & that kind of (yonder)  & あんな人 (that kind of person) \\
どんな & what kind of           & どんな映画？ (what kind of movie?) \\
\bottomrule
\end{tabular}
\end{center}
Do not add の: ($\times$) そんな\textbf{の}イベント $\rightarrow$
($\checkmark$) そんなイベント.
Also distinguish \textbf{そんな} (+ noun, ``such'') from \textbf{そんなに}
(+ adjective/verb, ``that much''): そんなに高い (``that expensive'').

\subsection{\texorpdfstring{知っています}{shitte imasu} (to know) --- class example}
To ask whether someone knows a person (or thing), use:
\begin{center}
  \textbf{[Person]を知っていますか。} \quad(\emph{[Person] wo shitte imasu ka})
\end{center}
``Do you know [Person]?'' --- e.g.\ 田中さんを知っていますか。

``To know'' is expressed as a \emph{state} using the て-form + います
(知る $\rightarrow$ 知\textbf{って}, small っ):
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Meaning & Japanese & Note \\
\midrule
Do you know? & 知っていますか & て-form + います \\
I know       & 知っています   & \\
I don't know & 知りません     & irregular --- \emph{not} 知っていません \\
\bottomrule
\end{tabular}
\end{center}
Note the irregular negative: ``I don't know'' is \textbf{知りません}
(\emph{shirimasen}), not 知っていません.

\subsection{Probability adverbs (\texorpdfstring{多分}{tabun} etc.)}
These adverbs signal how likely something is. Percentages are approximate and
subjective --- there is no official figure.
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Japanese & Romaji & $\approx$\% & Feel \\
\midrule
必ず           & kanarazu        & 100 & definitely, without fail \\
きっと         & kitto           & 80--90 & surely (often with conviction/hope) \\
\textbf{多分}  & \textbf{tabun}  & \textbf{70--80} & \textbf{probably} \\
恐らく         & osoraku         & 70 & likely (more formal / written) \\
もしかしたら   & moshika shitara & 30--50 & maybe, possibly \\
ひょっとしたら & hyotto shitara  & 20--40 & there's a slight chance \\
\bottomrule
\end{tabular}
\end{center}
\textbf{多分} ($\approx$70--80\%) means you \emph{expect} something but aren't
certain --- stronger than English ``maybe.'' It usually pairs with a conjecture
ending \textbf{でしょう / だろう} or \textbf{〜と思う}:
\begin{itemize}
  \item 多分行くと思います。 --- ``I think I'll probably go.''
  \item 多分雨が降るでしょう。 --- ``It will probably rain.''
\end{itemize}

\subsection{Particle \texorpdfstring{で}{de} (means/method) vs \texorpdfstring{は}{wa} (topic)}
\textbf{で} marks the \emph{means or method} used to do something --- ``by/using
X, [verb].'' It is easy to reach for \textbf{は} instead, since は is the first
particle most learners meet and starts to feel like a default, but は marks the
\emph{topic} (what the sentence is about), not the tool used.
\begin{itemize}
  \item カード\textbf{で}払えますか？ --- ``Can I pay \emph{by} card?''
        (\emph{kaado de haraemasu ka?})
  \item ペン\textbf{で}書きます。 --- ``I write \emph{with} a pen.''
  \item レストラン\textbf{で}食べます。 --- ``I eat \emph{at} a restaurant.''
        (で also marks \emph{where} an action happens.)
\end{itemize}
Compare: カード\textbf{は}払えますか would read as ``as for the card, can [it]
pay?'' --- topic-marked but missing the ``by means of'' meaning entirely.

\section{Phrases}

\subsection{Self-introduction}
\begin{quote}
こんにちは。私はティオです。どうぞよろしくお願いします。
\end{quote}
\emph{Konnichiwa. Watashi wa Tio desu. Douzo yoroshiku onegai shimasu.}\\
``Hello. I am Theoden. Pleased to meet you.''

\begin{itemize}
  \item 私 (\emph{watashi}) --- I
  \item は (\emph{wa}) --- topic particle (typed \texttt{ha})
  \item です (\emph{desu}) --- polite copula, ``am/is''
  \item どうぞよろしくお願いします --- set phrase, ``pleased to meet you''
\end{itemize}

\subsection{Studying phrases}
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Japanese & Romaji & English \\
\midrule
質問があります。 & shitsumon ga arimasu & I have a question. \\
これをノートに書いてください。 & kore wo nōto ni kaite kudasai & Please write this in the notes. \\
\bottomrule
\end{tabular}
\end{center}
Recall: 書いてください uses the て-form of 書く (\emph{kaku}), which is 書\textbf{いて}
(く $\rightarrow$ いて), not 書きて.

\section{駅 (At the Station)}

\subsection{Directions}

\textbf{Asking the way:}
\begin{quote}
駅までどうやって行きますか。
\end{quote}
\emph{Eki made dou yatte ikimasu ka.} --- ``How do I get to the station?''

\begin{itemize}
  \item 駅 (\emph{eki}) --- (train) station
  \item まで (\emph{made}) --- up to / as far as (destination)
  \item どうやって (\emph{dou yatte}) --- how / by what means
  \item 行きます (\emph{ikimasu}) --- go (ます-form)
  \item か --- question particle
\end{itemize}

\textbf{Answering (with 歩く, ``to walk''):}
\begin{quote}
歩いて行きます。\quad / \quad 駅まで歩きます。
\end{quote}
\emph{Aruite ikimasu. / Eki made arukimasu.} --- ``I go on foot. / I walk to
the station.''

Conjugations of 歩く (\emph{aruku}): ます-form \textbf{歩きます}
(\emph{arukimasu}); て-form \textbf{歩いて} (\emph{aruite}). Note how
\textbf{歩いて行きます} uses the て-form to connect 歩く to 行く --- the て-form
links verbs, while \textbf{ます} ends the sentence.


\section{Conversation}

\subsection{\texorpdfstring{あいづち}{aizuchi} (backchanneling)}
The most important listening behaviour in Japanese is \textbf{あいづち}
(\emph{aizuchi}) --- frequent short signals that show ``I'm listening, keep
going.'' They are used far more often than in English; staying silent can seem
inattentive or like disagreement.
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Japanese & Romaji & Feel \\
\midrule
はい / ええ / うん & hai / ee / un & ``yes / mm-hm'' (formal $\rightarrow$ casual) \\
そうですね         & sou desu ne   & ``that's right / I see'' \\
そうですか         & sou desu ka   & ``oh, is that so?'' \\
なるほど           & naruhodo      & ``I see / makes sense'' \\
へえ〜             & hee           & ``wow / really!'' (surprise) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{\texorpdfstring{オウム返し}{oumu-gaeshi} (echoing)}
\textbf{オウム返し} (literally ``parrot return'') is repeating a \emph{key word}
back --- often with rising intonation --- to show you are following, confirm
what you heard, or express surprise. Use a light touch (a word or two), not full
sentences.
\begin{quote}
A: 昨日、京都に行きました。 --- ``Yesterday I went to Kyoto.''\\
B: 京都に！いいですね。 --- ``To Kyoto! Nice!''
\end{quote}
Think あいづち first, オウム返し second: the goal is shared rhythm and presence,
not summarising everything back.

\section{Workbook}

\subsection{Translation practice (English $\rightarrow$ Japanese)}
Translate each sentence into Japanese. All use grammar and vocabulary from the
sections above. Write your answer in the space, then check the key below.

\begin{enumerate}
  \item Good evening. \vspace{1.5em}
  \item I am Tio. \vspace{1.5em}
  \item Nice to meet you. \vspace{1.5em}
  \item I study Japanese. \vspace{1.5em}
  \item I study Japanese with my girlfriend. \vspace{1.5em}
  \item Do you know Tanaka? \vspace{1.5em}
  \item I don't have time. \vspace{1.5em}
  \item Is there anything else? \vspace{1.5em}
  \item How do I get to the station? \vspace{1.5em}
  \item She is better at Japanese than me. \vspace{1.5em}
  \item This one was good. \vspace{1.5em}
  \item It is raining. \vspace{1.5em}
\end{enumerate}

\subsection{Answer key}
\begin{enumerate}
  \item こんばんは。
  \item 私はティオです。
  \item はじめまして。
  \item 日本語を勉強します。
  \item 彼女と日本語を勉強します。
  \item 田中さんを知っていますか。
  \item 時間がありません。
  \item 何か他にありますか。
  \item 駅までどうやって行きますか。
  \item 彼女は私より日本語が上手です。
  \item これはよかったです。
  \item 雨が降っています。
\end{enumerate}



\section{Prompts for Claude}


\section{Diary}


% --- Bibliography ---
\printbibliography

\end{CJK}
\end{document}
